---
title: "Predicting MIAC Baseball Game Outcomes"
subtitle: "Using Pre-Game Statistics to Forecast Winners"
author: "Ben Trutwin"
date: "May 2026"
output:
  ioslides_presentation:
    widescreen: true
    smaller: false

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center")
library(dplyr)
library(tidyr)
library(ggplot2)
library(glmnet)
library(randomForest)
library(pROC)
library(caret)
library(tibble)
library(knitr)
library(kableExtra)
library(gridExtra)
```

```{r load-data, cache=TRUE}
df <- read.csv("miac_modeling_2026.csv", stringsAsFactors = FALSE)

feature_cols <- c(
  "is_conference", "home_field",
  "home_rf_last5", "home_ra_last5", "home_win_last5", "home_rd_last5",
  "away_rf_last5", "away_ra_last5", "away_win_last5", "away_rd_last5",
  "home_rest_days", "away_rest_days",
  "home_sp_ip_last5", "away_sp_ip_last5",
  "home_era_proxy_last5", "away_era_proxy_last5",
  "home_era_season",  "home_whip_season",  "home_k_per9_season",
  "home_bb_per9_season", "home_opp_ba_season",
  "away_era_season",  "away_whip_season",  "away_k_per9_season",
  "away_bb_per9_season", "away_opp_ba_season",
  "home_fld_pct_season", "home_errors_season", "home_dp_season",
  "home_sba_season",  "home_csb_season",
  "away_fld_pct_season", "away_errors_season", "away_dp_season",
  "away_sba_season",  "away_csb_season"
)
target_col   <- "home_win"
feature_cols <- intersect(feature_cols, names(df))
model_df     <- df %>% select(all_of(c(feature_cols, target_col)))

na_rate  <- colMeans(is.na(model_df))
dropped  <- names(na_rate[na_rate > 0.50 & names(na_rate) != target_col])
model_df <- model_df %>% select(-all_of(dropped)) %>% drop_na()
model_df <- model_df %>%
  mutate(home_win = factor(home_win, levels = c(0, 1), labels = c("Loss", "Win")))

set.seed(42)
n         <- nrow(model_df)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))
train_df  <- model_df[train_idx, ]
test_df   <- model_df[-train_idx, ]

X_train <- model.matrix(home_win ~ . - 1, data = train_df)
y_train <- as.numeric(train_df$home_win) - 1
X_test  <- model.matrix(home_win ~ . - 1, data = test_df)
y_test  <- as.numeric(test_df$home_win) - 1
```

```{r fit-models, cache=TRUE}
evaluate_model <- function(probs, truth, threshold = 0.5) {
  preds   <- ifelse(probs >= threshold, 1, 0)
  acc     <- mean(preds == truth)
  roc_obj <- roc(truth, probs, quiet = TRUE)
  auc_val <- as.numeric(auc(roc_obj))
  cm <- confusionMatrix(
    factor(preds, levels = c(0,1), labels = c("Loss","Win")),
    factor(truth,  levels = c(0,1), labels = c("Loss","Win")),
    positive = "Win"
  )
  list(acc = acc, auc = auc_val, roc = roc_obj, cm = cm)
}

log_model <- glm(home_win ~ ., data = train_df, family = binomial)
log_probs <- predict(log_model, newdata = test_df, type = "response")
log_eval  <- evaluate_model(log_probs, y_test)

set.seed(42)
lasso_cv    <- cv.glmnet(X_train, y_train, family = "binomial", alpha = 1, nfolds = 10)
lasso_probs <- as.numeric(predict(lasso_cv, newx = X_test,
                                   s = "lambda.1se", type = "response"))
lasso_eval  <- evaluate_model(lasso_probs, y_test)

lasso_coef_df <- as.data.frame(as.matrix(coef(lasso_cv, s = "lambda.1se"))) %>%
  rownames_to_column("feature") %>%
  setNames(c("feature", "coefficient")) %>%
  filter(feature != "(Intercept)", coefficient != 0) %>%
  arrange(desc(abs(coefficient)))

set.seed(42)
rf_model <- randomForest(home_win ~ ., data = train_df,
                          ntree = 500, mtry = floor(sqrt(ncol(train_df) - 1)),
                          importance = TRUE)
rf_probs <- predict(rf_model, newdata = test_df, type = "prob")[, "Win"]
rf_eval  <- evaluate_model(rf_probs, y_test)

imp_df <- as.data.frame(importance(rf_model)) %>%
  rownames_to_column("feature") %>%
  arrange(desc(MeanDecreaseGini))
```


## What Is This Project?

<br>

> **Goal:** Build a model that predicts whether the **home team wins** a MIAC baseball game using only information available *before the first pitch.*

<br>

**Why does this matter?**

- Small-college baseball (DIII) is far less studied than professional or Division I analytics
- Home-field advantage at the DIII level is theoretically smaller, making it harder to isolate skill from noise
- The MIAC schedule is **short and weather-interrupted** — can rolling recent form capture team quality better than season averages?
- Practical value for coaches: lineup decisions, pitching deployment, tournament seed projections


## The Target Variable

```{r target-dist, fig.height=3.5, fig.width=6}
model_df %>%
  count(home_win) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = home_win, y = n, fill = home_win)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = paste0(n, "\n(", scales::percent(pct, accuracy = 1), ")")),
            vjust = -0.3, size = 5, fontface = "bold") +
  scale_fill_manual(values = c("Loss" = "#d9534f", "Win" = "#5cb85c")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Home Team Outcomes — 2026 MIAC Baseball",
       x = "Outcome", y = "Number of Games") +
  theme_minimal(base_size = 14)
```

The home team won **57%** of completed games. A naive "always predict home win" baseline achieves only 57% accuracy — the model must do better.


## Variables Used (1/2) — Game Context & Recent Form

| Group | Variables | Why Important |
|---|---|---|
| **Game context** | `is_conference`, `home_field` | Conference dynamics; crowd/travel advantage |
| **Rolling form (last 5)** | `rf_last5`, `ra_last5`, `win_last5`, `rd_last5` (home & away) | Short-term momentum and opponent quality |
| **Pitcher load (last 5)** | `sp_ip_last5`, `era_proxy_last5` (home & away) | Starter availability and recent effectiveness |


## Variables Used (2/2) — Season Statistics

| Group | Variables | Why Important |
|---|---|---|
| **Away pitching (season)** | `era_season`, `whip_season`, `k_per9_season`, `bb_per9_season`, `opp_ba_season` | Stable baseline for team pitching quality |
| **Away fielding (season)** | `fld_pct_season`, `errors_season`, `dp_season`, `sba_season`, `csb_season` | Defensive efficiency |

> **Note:** Home season-level stats were **dropped** (>50% missing values) — a data asymmetry discussed later.


## Where Did the Data Come From?

<br>

**Source:** [miacathletics.com](https://miacathletics.com) — the official MIAC record-keeper

<br>

**How it was collected:**

1. Custom R web scraper using `rvest` + `chromote` (headless Chrome for JavaScript-rendered pages)
2. Scraped the **full composite schedule** (all months, Feb–May 2026)
3. Scraped **every box score** for every completed game
4. Scraped **team-level season stats** (pitching, batting, fielding)
5. Engineered **rolling (last-5-game) features** using a sliding window — *no future data leakage*

<br>

**Result:** 224 completed games scraped → **138 usable rows** after complete-case filtering


## Signal #1 — Recent Wins Matter

```{r eda-win-last5, fig.height=3.6, fig.width=9}
p1 <- ggplot(model_df, aes(x = home_win, y = home_win_last5, fill = home_win)) +
  geom_boxplot(alpha = 0.75, show.legend = FALSE, outlier.shape = 21) +
  scale_fill_manual(values = c("Loss" = "#d9534f", "Win" = "#5cb85c")) +
  labs(title = "Home Wins in Last 5 Games", x = "Outcome", y = "Wins (Last 5)") +
  theme_minimal(base_size = 13)

p2 <- ggplot(model_df, aes(x = home_win, y = home_rd_last5, fill = home_win)) +
  geom_boxplot(alpha = 0.75, show.legend = FALSE, outlier.shape = 21) +
  scale_fill_manual(values = c("Loss" = "#d9534f", "Win" = "#5cb85c")) +
  labs(title = "Home Run Differential (Last 5)", x = "Outcome", y = "Run Diff (Last 5)") +
  theme_minimal(base_size = 13)

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

Teams with **more wins** and a **positive run differential** over their last 5 games are far more likely to win their next home game. Short-term momentum is the dominant signal.


## Signal #2 — Away ERA is Weaker Than Expected

```{r eda-away-era, fig.height=3.6, fig.width=6}
ggplot(model_df, aes(x = home_win, y = away_era_season, fill = home_win)) +
  geom_boxplot(alpha = 0.75, show.legend = FALSE, outlier.shape = 21) +
  scale_fill_manual(values = c("Loss" = "#d9534f", "Win" = "#5cb85c")) +
  labs(title = "Away Team Season ERA vs. Home Team Outcome",
       x = "Home Team Outcome", y = "Away Team Season ERA") +
  theme_minimal(base_size = 14)
```

Away ERA shows only a **modest** separation — the MIAC is an offensive conference, and unearned runs (not captured in ERA) may shift outcomes without showing up in this stat.



## Design Decisions (1/2)

**Preventing data leakage**
> All features computed strictly from information *prior* to each game. Scores and run differentials only appear in rolling windows ending *before* the target game.

**Handling missing data**
> Home-team season-level stats exceeded **50% missing** and were dropped rather than imputed — imputing over half a column adds noise, not signal.

**Train/test split:** 80/20 random split — 110 training observations, 28 test observations (seed = 42).


## Design Decisions (2/2) — Three Model Types

| Model | Why |
|---|---|
| **Logistic Regression** | Interpretable baseline; well-suited to binary classification |
| **LASSO** | L1 regularization performs automatic feature selection; combats overfitting on small data |
| **Random Forest** | Captures nonlinear interactions; robust ensemble approach |

> Best Subset Selection returned a degenerate solution (0 selected variables) on this small sample — excluded from comparison.


## LASSO: Automatic Feature Selection

```{r lasso-cv-slide, fig.height=3.8, fig.width=6}
plot(lasso_cv, main = "LASSO 10-Fold CV — Binomial Deviance vs. log(λ)")
```

Using the **1-SE rule**, LASSO shrunk all but **3 predictors** to zero, avoiding overfitting.


## LASSO: The 3 Predictors That Survived

```{r lasso-coef-slide, fig.height=3.5, fig.width=7}
ggplot(lasso_coef_df, aes(x = reorder(feature, coefficient), y = coefficient,
                           fill = coefficient > 0)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "#5cb85c", "FALSE" = "#d9534f")) +
  coord_flip() +
  labs(title = "LASSO Selected Coefficients (λ = 1-SE)",
       x = NULL, y = "Coefficient") +
  theme_minimal(base_size = 14)
```

1. `home_win_last5` — Home team recent wins *(positive → more likely to win)*
2. `away_win_last5` — Away team recent wins *(negative → harder to beat good visitors)*
3. `away_rd_last5` — Away team run differential *(negative → strong visitor is a threat)*


## Random Forest: Variable Importance

```{r rf-imp-slide, fig.height=4, fig.width=8}
ggplot(imp_df[1:12, ], aes(x = reorder(feature, MeanDecreaseGini),
                             y = MeanDecreaseGini)) +
  geom_col(fill = "forestgreen") +
  coord_flip() +
  labs(title = "Top 12 Features — Random Forest (Mean Decrease Gini)",
       x = NULL, y = "Mean Decrease Gini") +
  theme_minimal(base_size = 13)
```

The Random Forest agrees: **recent run differential and wins** dominate. Season-level stats contribute far less — short-term momentum is the primary signal.


## Model Performance Comparison

```{r perf-table-slide}
summary_tbl <- tibble(
  Model       = c("Logistic Regression", "LASSO", "Random Forest"),
  Accuracy    = c(log_eval$acc,  lasso_eval$acc,  rf_eval$acc),
  AUC         = c(log_eval$auc,  lasso_eval$auc,  rf_eval$auc),
  Sensitivity = c(log_eval$cm$byClass["Sensitivity"],
                  lasso_eval$cm$byClass["Sensitivity"],
                  rf_eval$cm$byClass["Sensitivity"]),
  Specificity = c(log_eval$cm$byClass["Specificity"],
                  lasso_eval$cm$byClass["Specificity"],
                  rf_eval$cm$byClass["Specificity"])
) %>% mutate(across(where(is.numeric), ~ round(., 3)))

kable(summary_tbl, align = c("l","c","c","c","c")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                font_size = 18, full_width = TRUE) %>%
  row_spec(which.max(summary_tbl$AUC), bold = TRUE, background = "#d4edda")
```

**LASSO wins on every metric** — 85.7% accuracy and AUC = 0.887 with only 3 features. The full logistic regression (26 features) is clearly overfitting the 110-observation training set.


## ROC Curve — Logistic Regression

```{r roc-logistic-slide, fig.height=4, fig.width=6}
plot(log_eval$roc, col = "blue", lwd = 2.5,
     main = paste0("Logistic Regression  |  AUC = ", round(auc(log_eval$roc), 3)))
abline(a = 0, b = 1, lty = 2, col = "gray60")
```

The curve shows how well the model ranks wins vs. losses at every threshold. AUC = 1.0 is perfect; AUC = 0.5 is a coin flip.


## ROC Curve — LASSO

```{r roc-lasso-slide, fig.height=4, fig.width=6}
plot(lasso_eval$roc, col = "red", lwd = 2.5,
     main = paste0("LASSO  |  AUC = ", round(auc(lasso_eval$roc), 3)))
abline(a = 0, b = 1, lty = 2, col = "gray60")
```

LASSO achieves the highest AUC with only **3 features**. The curve hugs the top-left corner — strong discrimination between wins and losses.


## ROC Curve — Random Forest

```{r roc-rf-slide, fig.height=4, fig.width=6}
plot(rf_eval$roc, col = "darkgreen", lwd = 2.5,
     main = paste0("Random Forest  |  AUC = ", round(auc(rf_eval$roc), 3)))
abline(a = 0, b = 1, lty = 2, col = "gray60")
```

The Random Forest shows reasonable discrimination but trails LASSO, likely due to the small 28-game test set amplifying ensemble variance.


## Why Did LASSO Win? (Bias-Variance Tradeoff)

<br>

| Factor | What Happened |
|---|---|
| **Small dataset** (110 training obs.) | Complex models memorize noise; regularization protects against this |
| **26 predictors, 110 rows** | Logistic regression is underdetermined — coefficients are unreliable |
| **LASSO's L1 penalty** | Forces most coefficients to zero, keeping only the 3 strongest signals |
| **Random Forest variance** | Ensemble variance is high with a 28-observation test set |

<br>

> **Key insight:** In this DIII short-season context, *recent form* is far more informative than long-run season statistics.


## What Could Make This Better?

**1. Multi-season data** — Training on 2023–2026 would grow the sample from ~138 to 500+ observations.

**2. Named starting pitcher** — Day-of-game starter ERA/WHIP is the strongest pre-game baseball signal; couldn't get a consistent match at this dataset size.

**3. Head-to-head history** — Recent series results and rivalry effects carry signal beyond aggregate rolling stats.

**4. Weather & field conditions** — Temperature and wind affect scoring but are not currently included.

**5. Strength of schedule** — DIII talent gaps are enormous; no opponent-quality metric was available in this dataset.


## Key Takeaways

- **MIAC baseball outcomes are predictable** at a meaningfully above-chance level using only pre-game public data
- **LASSO with 3 features** outperformed models with 26 features — a textbook case of regularization beating complexity on small data
- **Short-term momentum** (wins and run differential over the last 5 games) is the dominant signal; season-long stats add little beyond it
- **Home-field advantage exists** (57% home win rate), but it is not the main driver of predictions
- **Primary limitation:** 138 usable games is a small sample — multi-season data is the single highest-impact improvement

> *"In short-season DIII baseball, what you did last week matters more than what you did all year."*


## Thank You

<br>

**Data source:** MIAC Athletics — [miacathletics.com](https://miacathletics.com)

**Tools used:** R, `rvest`, `chromote`, `glmnet`, `randomForest`, `pROC`, `caret`, `ggplot2`

**Files:**
- `MIAC_Full_Scrape.Rmd` — Web scraper
- `Feature_Engineering_Final.Rmd` — Feature construction
- `modeling.Rmd` — Model fitting
- `Final_Report.rmd` — Full written report

<br>

> Questions?
