---
title: "NCAA Mini-Project - Part 1"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# put your library statements here
library(tidyverse)
library(caret)

# import the data
mens_data <- read_csv("https://raw.githubusercontent.com/gmtanner-cord/March-Madness/refs/heads/main/training_data/ncaa_mbb_2001-2025_full.csv")
womens_data <- read_csv("https://raw.githubusercontent.com/gmtanner-cord/March-Madness/refs/heads/main/training_data/ncaa_wbb_2001-2025_full.csv")

# choose whether you want to predict the mens or womens tournament
ncaa_data <- mens_data
ncaa_data <- ncaa_data %>% mutate(winner = factor(winner, levels = c("team1","team2")))
```

## The Data

Data was collected from
[Sports-Reference](https://www.sports-reference.com/cbb/seasons/2019-school-stats.html)
for the NCAA Men's and Women's basketball tournaments from 2001-2025 (no
tourney in 2020). The data was collected by official scorers at the
college basketball games. One flaw in the data is that the season stats
also include the performance in the NCAA tournament. This does affect
its predictive ability for trying to predict the results of a future
tournament, but it would be a lot more work to fix the data.

There are 134 variables in the dataset (2 response variables, 2
variables about for the round and year, and 65 for each team with ".1"
appended for team 1 and ".2" appended for team 2). Here are some of the
predictor variables:

-   seed: Number 1-16. Determined by a 10-person committee. Lower seeds
    are better.
-   win_pct: Win Percentage (total, home, away, conf). Calculated as
    wins over games played.
-   srs: Simple Ratings System. A rating that takes into account average
    point differential and strength of schedule. The rating is
    denominated in points above/below average, where zero is average.
    Non-Division I games are excluded from the ratings.
-   sos: Strength of Schedule. A rating of strength of schedule. The
    rating is denominated in points above/below average, where zero is
    average. sos is srs minus the team's average point differential.
-   pts_pg: Points per Game. Number of points scored divided by number
    of games.
-   opp_pts_pg: Opponents Points per Games. Number of points scored by
    opponents divided by number of games.
-   pace: An estimate of number of possessions per 40 minutes.
-   off_rtg: Offensive rating. Points scored per 100 possessions.
-   fg_pct, fg3_pct, ft_pct: Field Goals (2 points), 3-Point Field
    Goals, Free Throw (1 point) percentages. Number scored divided by
    number attempted.
-   ts_pct: True shooting percentage. A measure of shooting efficiency
    that takes into account 2-point field goals, 3-point field goals,
    and free throws.
-   trb_pct: Total Rebound Percentage. An estimate of the percentage of
    available rebounds a player grabbed while they were on the floor.
-   ast_pct: An estimate of the percentage of the teams field goals that
    a player assisted.
-   stl_pct: Steal Percentage. An estimate of the percentage of opponent
    possessions that end with a steal.
-   blk_pct: Block Percentage. An estimate of the percentage of opponent
    two-point field goal attempts blocked.
-   efg_pct: Effective Field Goal Percentage; this statistic adjusts for
    the fact that a 3-point field goal is worth one more point than a
    2-point field goal.
-   tov_pct: Turnover Percentage; an estimate of turnovers per 100
    plays.
-   orb_pct: Offensive Rebound Percentage; an estimate of the percentage
    of available offensive rebounds.
-   ft_rate: Free Throws Made Per Field Goal Attempt
-   \*\_pg: Counting Stats per Game, including field goals (fg), 3-Point
    Field Goals (fg3), Free Throw (ft), offensive rebounds (orb), total
    rebounds (trb), assists (ast), steals (stl), blocks (blk), turnovers
    (tov), and personal fouls (pf).
-   \*\_opp: Variables that denote the amount of that stat for the
    team's opponents.
-   conf: Conference
-   mp_bench_pct, pts_bench_pct: Percent of minutes played/points scored
    by the bench players (players ranked 6th or more by minutes played)
-   height_avg, height_max: Average and max height of players of the top
    6 players (by minutes played) on the roster.
-   exp_avg: Average experience (Freshman = 1 year, ..., Senior = 4
    years) of the top 6 players (by minutes played) on the roster.
-   rsci_pct: Percentage of players ranked by RSCI during their senior
    year of high school of the top 6 players (by minutes played) on the
    roster.
-   min_return, pts_return: Percent of minutes played/points scored
    returned from previous year's squad.

There are two response variables:

-   winner: has values "team1" and "team2" for which team won the game.
-   point_diff: difference of scores team1-team2 in the game

The dataset has been mirrored so that each game shows up twice, once
with each of the teams as team 1. This was done to alleviate the bias in
the data towards team 1 winning (typically the better seed was listed
first as team 1 in the early rounds of the bracket) and to make your
classifiers robust to the order of the teams (you wouldn't want to model
to produce different predictions based on which team was listed as team
1). Just be aware that doing this will underestimate the uncertainty in
the parameter estimates (your p-values will be smaller than they should
be).


## Trapezoid of Excellence

The "Trapezoid of Excellence" is a visual predictor that plots each team's **pace** (estimated possessions per 40 minutes) against their **net rating** (offensive efficiency minus defensive efficiency, adjusted for opponent strength). The key insight is that teams at unusual paces (very slow or very fast) must have an exceptional net rating to be elite, while teams at an average pace can succeed with a somewhat lower net rating. The trapezoid's lower boundary therefore dips at average pace (~65) and rises toward the extremes.

**Prediction rules:**
1. If **one team is in the trapezoid and the other is not**, the team inside the trapezoid wins.
2. If **both teams are outside**, the team *closer* (smaller distance) to the trapezoid boundary wins.
3. If **both teams are inside**, the team with the *higher net rating* wins.

### Step 1: Compute Net Rating

Net rating = offensive rating (points scored per 100 possessions) minus defensive rating (points allowed per 100 possessions). `off_rtg` is provided directly. Defensive rating is estimated as opponent points per game scaled by possessions.

```{r trapezoid-net-rating}
# Defensive rating: opponent points per game scaled to per-100-possessions
# net rating = off_rtg (pts scored per 100 poss) minus def_rtg (pts allowed per 100 poss)
ncaa_data <- ncaa_data %>%
  mutate(
    def_rtg.1 = (pts_pg_opp.1 / pace.1) * 100,
    def_rtg.2 = (pts_pg_opp.2 / pace.2) * 100,
    net_rtg.1 = off_rtg.1 - def_rtg.1,
    net_rtg.2 = off_rtg.2 - def_rtg.2
  )

# Sanity check: net rating distribution should center near 0
summary(ncaa_data$net_rtg.1)
```

### Step 2: Define the Trapezoid Boundary

The lower boundary of the trapezoid is a V-shaped function of pace. At the center pace (`trap_center_pace`), the minimum net rating to qualify is `trap_min_net_rtg`. For every unit a team's pace deviates from center, `trap_slope` additional net rating points are required.

These parameters are estimated by eye from the original Trapezoid of Excellence plot (pace ~59–73, lowest threshold around net rating 22 at pace 65).

```{r trapezoid-define}
# --- Trapezoid parameters (tunable) ---
# The boundary is FLAT between trap_lo and trap_hi, then rises asymmetrically:
#   - Left wall is steep (slow-pace teams must be elite to qualify)
#   - Right wall is gentler (fast-pace teams get slightly more leeway)
# Calibrated so that at pace ~59, threshold ~35 (left), and at pace ~76, threshold ~26 (right).
trap_lo          <- 62.5  # left corner of flat bottom
trap_hi          <- 72.0  # right corner of flat bottom
trap_min_net_rtg <- 22.0  # net rating threshold along the flat bottom
trap_left_slope  <- 4.5   # net rating rise per unit pace BELOW trap_lo
trap_right_slope <- 1.5   # net rating rise per unit pace ABOVE trap_hi

# Lower boundary: flat between trap_lo and trap_hi, rising asymmetrically outside
trap_lower <- function(pace) {
  trap_min_net_rtg +
    trap_left_slope  * pmax(0, trap_lo - pace) +
    trap_right_slope * pmax(0, pace - trap_hi)
}

# Signed distance to the trapezoid boundary:
#   negative (or zero) = INSIDE the trapezoid (net_rtg meets or exceeds threshold)
#   positive           = OUTSIDE the trapezoid (net_rtg falls short by this much)
dist_to_trap <- function(pace, net_rtg) {
  trap_lower(pace) - net_rtg
}
```

### Step 3: Engineer Trapezoid Features

```{r trapezoid-features}
# Ensure net_rtg columns exist (re-computes if Step 1 wasn't run yet)
if (!"net_rtg.1" %in% names(ncaa_data)) {
  ncaa_data <- ncaa_data %>%
    mutate(
      def_rtg.1 = (pts_pg_opp.1 / pace.1) * 100,
      def_rtg.2 = (pts_pg_opp.2 / pace.2) * 100,
      net_rtg.1 = off_rtg.1 - def_rtg.1,
      net_rtg.2 = off_rtg.2 - def_rtg.2
    )
}

ncaa_data <- ncaa_data %>%
  mutate(
    # Boolean: is the team inside the trapezoid?
    in_trap.1 = dist_to_trap(pace.1, net_rtg.1) <= 0,
    in_trap.2 = dist_to_trap(pace.2, net_rtg.2) <= 0,

    # Signed distance (negative = inside, positive = outside)
    dist_trap.1 = dist_to_trap(pace.1, net_rtg.1),
    dist_trap.2 = dist_to_trap(pace.2, net_rtg.2),

    # Unified "trapezoid advantage" for team 1 over team 2.
    # Positive values favour team 1; negative values favour team 2.
    # Rule 1: team1 in, team2 out  -> advantage proportional to team2's exclusion distance
    # Rule 2: team1 out, team2 in  -> negative advantage proportional to team1's exclusion distance
    # Rule 3: both in              -> net rating difference decides
    # Rule 4: both out             -> the team closer to the boundary has the advantage
    trap_advantage = case_when(
      in_trap.1 & !in_trap.2  ~  dist_trap.2,                  # Rule 1
      !in_trap.1 & in_trap.2  ~ -dist_trap.1,                  # Rule 2
      in_trap.1 & in_trap.2   ~  net_rtg.1 - net_rtg.2,        # Rule 3
      TRUE                    ~  dist_trap.2 - dist_trap.1      # Rule 4
    )
  )

# How many games have at least one team inside the trapezoid?
ncaa_data %>%
  summarise(
    pct_team1_in = mean(in_trap.1, na.rm = TRUE),
    pct_team2_in = mean(in_trap.2, na.rm = TRUE),
    pct_both_in  = mean(in_trap.1 & in_trap.2, na.rm = TRUE),
    pct_both_out = mean(!in_trap.1 & !in_trap.2, na.rm = TRUE)
  )
```

### Step 4: Visualise the Trapezoid

```{r trapezoid-viz, fig.width=8, fig.height=5}
# Extend pace grid to full axis range so both diagonal walls reach the edges
pace_grid <- seq(57, 78, by = 0.1)
trap_df <- data.frame(
  pace      = pace_grid,
  threshold = trap_lower(pace_grid)
)

# Sample one observation per game (avoid double-counting the mirrored rows)
viz_data <- head(ncaa_data, nrow(ncaa_data) / 2) %>%
  select(pace = pace.1, net_rtg = net_rtg.1, in_trap = in_trap.1) %>%
  filter(!is.na(pace), !is.na(net_rtg))

ggplot() +
  # Shade the "inside" region
  geom_ribbon(
    data = trap_df,
    aes(x = pace, ymin = threshold, ymax = 35),
    fill = "lightgreen", alpha = 0.25
  ) +
  # Draw the boundary line
  geom_line(
    data = trap_df,
    aes(x = pace, y = threshold),
    color = "darkgreen", linewidth = 1.2, linetype = "dashed"
  ) +
  # Plot teams, coloured by whether they're inside
  geom_point(
    data = viz_data,
    aes(x = pace, y = net_rtg, color = in_trap),
    alpha = 0.4, size = 1.5
  ) +
  scale_color_manual(
    values = c("TRUE" = "forestgreen", "FALSE" = "tomato"),
    labels = c("TRUE" = "In Trapezoid", "FALSE" = "Outside"),
    name   = ""
  ) +
  annotate("text", x = 65, y = trap_lower(65) - 1.5,
           label = "Trapezoid lower boundary", color = "darkgreen", size = 3.5) +
  labs(
    title = "Trapezoid of Excellence",
    subtitle = paste0("Flat zone pace ", trap_lo, "–", trap_hi,
                      ", min net rating = ", trap_min_net_rtg,
                      ", left slope = ", trap_left_slope,
                      ", right slope = ", trap_right_slope),
    x = "Pace (possessions per 40 min)",
    y = "Net Rating (pts per 100 poss)"
  ) +
  coord_cartesian(ylim = c(NA, 35)) +
  theme_bw()
```

### Step 5: Naive Trapezoid Rule Accuracy

Before fitting any ML model, check how well the raw `trap_advantage` sign predicts the winner.

```{r trapezoid-naive-accuracy}
ncaa_data_clean <- ncaa_data %>% filter(!is.na(trap_advantage))

naive_preds <- ifelse(ncaa_data_clean$trap_advantage >= 0, "team1", "team2")
naive_acc   <- mean(naive_preds == ncaa_data_clean$winner, na.rm = TRUE)
cat("Naive trapezoid rule accuracy:", round(naive_acc * 100, 1), "%\n")
```

### Step 6: Regression Model Using Trapezoid Advantage

`trap_advantage` is used as the primary predictor of point differential. The sign and magnitude directly encode the trapezoid rules, so a simple linear model tests whether the theory holds up statistically.

```{r trapezoid-reg-model}
# Re-build train/test respecting symmetry
middle <- nrow(ncaa_data_clean) / 2
top_half_t  <- head(ncaa_data_clean, middle)
bot_half_t  <- tail(ncaa_data_clean, middle)
set.seed(42)
idx_t       <- createDataPartition(top_half_t$winner, p = 0.20, list = FALSE)
train_trap  <- rbind(top_half_t[-idx_t, ], bot_half_t[-idx_t, ])
test_trap   <- rbind(top_half_t[ idx_t, ], bot_half_t[ idx_t, ])

# Regression: predict point_diff
trap_lm <- lm(point_diff ~ trap_advantage, data = train_trap)
summary(trap_lm)

yhat_reg <- predict(trap_lm, newdata = test_trap)
rmse_trap <- sqrt(mean((test_trap$point_diff - yhat_reg)^2, na.rm = TRUE))
r2_trap   <- cor(test_trap$point_diff, yhat_reg, use = "complete.obs")^2
cat("Test RMSE:", round(rmse_trap, 3), "\n")
cat("Test R²  :", round(r2_trap,   3), "\n")

# Visualise fit
ggplot(train_trap, aes(x = trap_advantage, y = point_diff)) +
  geom_jitter(alpha = 0.25, width = 0.3) +
  geom_smooth(method = "lm", color = "steelblue") +
  labs(
    title = "Point Differential vs. Trapezoid Advantage",
    x     = "Trapezoid Advantage (team 1 over team 2)",
    y     = "Actual Point Differential"
  ) +
  theme_bw()
```

### Step 7: Classification Model Using Trapezoid Advantage

```{r trapezoid-class-model}
# Logistic regression: predict winner
trap_glm <- glm(winner ~ trap_advantage, data = train_trap,
                family = binomial(link = "logit"))
summary(trap_glm)

# Predicted probabilities and class labels on test set
prob_trap  <- predict(trap_glm, newdata = test_trap, type = "response")
pred_class <- factor(ifelse(prob_trap >= 0.5, "team1", "team2"),
                     levels = c("team1", "team2"))

cm_trap <- confusionMatrix(pred_class, test_trap$winner, positive = "team1")
print(cm_trap)

# ROC curve
library(pROC)
roc_trap <- roc(test_trap$winner, prob_trap, levels = c("team2", "team1"))
cat("AUC:", round(auc(roc_trap), 3), "\n")
plot(roc_trap, main = "ROC – Trapezoid Logistic Model", print.auc = TRUE)

# Visualise probability curve
adv_grid <- data.frame(trap_advantage = seq(-20, 20, 0.5))
adv_grid$p_team1 <- predict(trap_glm, newdata = adv_grid, type = "response")
ggplot(adv_grid, aes(x = trap_advantage, y = p_team1)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray40") +
  labs(
    title = "Predicted P(Team 1 Wins) vs. Trapezoid Advantage",
    x     = "Trapezoid Advantage",
    y     = "P(team1 wins)"
  ) +
  theme_bw()
```

### Step 8: Apply to Specific Match-ups

```{r trapezoid-matchups}
# Define match-up rows
mens_matchups <- mens_data[c(1509,1511,2613),]

# Enrich the pre-loaded match-up rows with trapezoid features
enrich_matchups <- function(df) {
  df %>%
    mutate(
      def_rtg.1      = (pts_pg_opp.1 / pace.1) * 100,
      def_rtg.2      = (pts_pg_opp.2 / pace.2) * 100,
      net_rtg.1      = off_rtg.1 - def_rtg.1,
      net_rtg.2      = off_rtg.2 - def_rtg.2,
      in_trap.1      = dist_to_trap(pace.1, net_rtg.1) <= 0,
      in_trap.2      = dist_to_trap(pace.2, net_rtg.2) <= 0,
      dist_trap.1    = dist_to_trap(pace.1, net_rtg.1),
      dist_trap.2    = dist_to_trap(pace.2, net_rtg.2),
      trap_advantage = case_when(
        in_trap.1 & !in_trap.2  ~  dist_trap.2,
        !in_trap.1 & in_trap.2  ~ -dist_trap.1,
        in_trap.1 & in_trap.2   ~  net_rtg.1 - net_rtg.2,
        TRUE                    ~  dist_trap.2 - dist_trap.1
      )
    )
}

womens_matchups <- womens_data[c(1446,1503,1509),]

mens_matchups_trap   <- enrich_matchups(mens_matchups)
womens_matchups_trap <- enrich_matchups(womens_matchups)

# Regression prediction (point differential)
mens_matchups_trap$pred_point_diff   <- predict(trap_lm,  newdata = mens_matchups_trap)
womens_matchups_trap$pred_point_diff <- predict(trap_lm,  newdata = womens_matchups_trap)
# Classification prediction (win probability)
mens_matchups_trap$p_team1_wins   <- predict(trap_glm, newdata = mens_matchups_trap,   type = "response")
womens_matchups_trap$p_team1_wins <- predict(trap_glm, newdata = womens_matchups_trap, type = "response")

cat("--- Men's Matchups ---\n")
mens_matchups_trap %>%
  select(trap_advantage, in_trap.1, in_trap.2,
         net_rtg.1, net_rtg.2, pace.1, pace.2,
         pred_point_diff, p_team1_wins)

cat("--- Women's Matchups ---\n")
womens_matchups_trap %>%
  select(trap_advantage, in_trap.1, in_trap.2,
         net_rtg.1, net_rtg.2, pace.1, pace.2,
         pred_point_diff, p_team1_wins)
```

## Feature Engineering & Data Preparation

Before building the remaining models, I'm making difference variables for each predictor and creating a single shared train/test split. Using the same split across all models ensures performance comparisons are the same.

The six predictors chosen are:
- `trap_advantage` — trapezoid-based composite score (already computed above)
- `srs_diff` — Simple Rating System difference; captures overall team quality adjusted for schedule
- `seed_diff` — seed difference; encodes the committee's seeding judgment
- `ts_pct_diff` — True Shooting % difference; comprehensive shooting efficiency
- `tov_pct_diff` — Turnover % difference; ball control (note: higher tov_pct is *worse*, so positive diff means team 1 turns it over more)
- `trb_pct_diff` — Total Rebound % difference; rebounding dominance

These six cover the main pillars of basketball performance (composite efficiency, schedule strength, seeding, shooting, ball security, rebounding) with minimal overlap.

```{r feature-engineering}
# Engineer difference variables (team1 minus team2 for each stat)
ncaa_data <- ncaa_data %>%
  mutate(
    srs_diff     = srs.1     - srs.2,
    seed_diff    = seed.1    - seed.2, 
    ts_pct_diff  = ts_pct.1  - ts_pct.2,
    tov_pct_diff = tov_pct.1 - tov_pct.2,
    trb_pct_diff = trb_pct.1 - trb_pct.2
  )

# Quick correlation check among predictors (should be low)
ncaa_data %>%
  select(trap_advantage, srs_diff, seed_diff, ts_pct_diff, tov_pct_diff, trb_pct_diff) %>%
  cor(use = "complete.obs") %>%
  round(2)
```

```{r shared-split}
# Keep only rows where all 6 predictors and both responses are complete
model_vars <- c("trap_advantage", "srs_diff", "seed_diff",
                "ts_pct_diff", "tov_pct_diff", "trb_pct_diff",
                "point_diff", "winner")

ncaa_model <- ncaa_data %>% # Keeps only 8 cols I need and drops any row where any of the cols is NA
  select(all_of(model_vars)) %>%
  filter(complete.cases(.))

cat("Rows available for modeling:", nrow(ncaa_model), "\n")

# Symmetry-preserving 80/20 train/test split - I don't fully understand why I need to use both halves, but I asked Copilot on training and it was very strongly suggesting to use both for more training and for the model to learn symmetry. 
set.seed(42)
mid        <- nrow(ncaa_model) / 2
top_half   <- head(ncaa_model, mid)
bot_half   <- tail(ncaa_model, mid)
test_idx   <- createDataPartition(top_half$winner, p = 0.20, list = FALSE)

train_set  <- rbind(top_half[-test_idx, ], bot_half[-test_idx, ])
test_set   <- rbind(top_half[ test_idx, ], bot_half[ test_idx, ])

cat("Training rows:", nrow(train_set), " | Test rows:", nrow(test_set), "\n")
```

## Best Subset Selection (BeSS)

Best Subset Selection fits every possible combination of the 6 predictors and identifies the single best model of each size (1 variable, 2 variables, …, 6 variables). We then use AIC and BIC to choose the optimal number of variables to include. BIC penalizes complexity more heavily than AIC, so it tends to choose smaller models.

### BeSS Regression

```{r bess-reg-fit}
library(BeSS)

# bess() needs a matrix of inputs, not a formula — use model.matrix() to create it
x_train_reg <- model.matrix(point_diff ~ trap_advantage + srs_diff + seed_diff +
                               ts_pct_diff + tov_pct_diff + trb_pct_diff,
                             data = train_set)
x_test_reg  <- model.matrix(point_diff ~ trap_advantage + srs_diff + seed_diff +
                               ts_pct_diff + tov_pct_diff + trb_pct_diff,
                             data = test_set)

# Fit all model sizes sequentially (family = "gaussian" for linear regression)
bess_reg <- bess(x = x_train_reg, y = train_set$point_diff,
                 family = "gaussian", method = "sequential")

bess_reg_summary <- data.frame(print(bess_reg))

# Plot AIC and BIC vs model size to choose best number of variables
ggplot(bess_reg_summary) +
  geom_line(aes(x = Df, y = AIC), linetype = "solid") +
  geom_line(aes(x = Df, y = BIC), linetype = "dashed") +
  annotate("point", x = which.min(bess_reg_summary$AIC), y = min(bess_reg_summary$AIC)) +
  annotate("point", x = which.min(bess_reg_summary$BIC), y = min(bess_reg_summary$BIC)) +
  labs(title = "BeSS Regression: AIC (solid) and BIC (dashed) vs Model Size",
       x = "Model Size", y = "AIC (solid) / BIC (dashed)")

# Best model size according to BIC
best_size_reg <- which.min(bess_reg_summary$BIC)
cat("Best model size (BIC):", best_size_reg, "\n")

# Non-zero coefficients in the BIC-selected model
BIC_coef_reg <- coef(bess_reg)[, best_size_reg]
cat("Variables included:\n")
print(BIC_coef_reg[BIC_coef_reg != 0])
```

```{r bess-reg-predict}
# Test set performance using bess predict with type = "BIC"
yhat_bess_reg <- predict(bess_reg, newdata = x_test_reg, type = "BIC")
rmse_bess_reg <- sqrt(mean((test_set$point_diff - yhat_bess_reg)^2))
r2_bess_reg   <- cor(test_set$point_diff, yhat_bess_reg)^2
cat("BeSS Regression — Test RMSE:", round(rmse_bess_reg, 3),
    " | Test R²:", round(r2_bess_reg, 3), "\n")

# Refit as plain lm() with selected variables for visualization and impact analysis
best_coefs_reg <- names(BIC_coef_reg[BIC_coef_reg != 0])
best_coefs_reg <- best_coefs_reg[!grepl("intercept", best_coefs_reg, ignore.case = TRUE)] # needed to use grep1 to catch capital or lowercase intercept
# refits a standard linear model using just vars chosen by BeSS to use for future plots and interpretation
bess_lm <- lm(as.formula(paste("point_diff ~", paste(best_coefs_reg, collapse = " + "))),
              data = train_set)
summary(bess_lm)
```

```{r bess-reg-viz}
# Means of selected predictors — used to hold non-varying predictors constant
pred_means <- colMeans(train_set[, best_coefs_reg, drop = FALSE])

# Plot 1: point_diff vs srs_diff (seed_diff held at its mean)
grid_srs <- data.frame(
  srs_diff  = seq(quantile(train_set$srs_diff,  0.01),
                  quantile(train_set$srs_diff,  0.99), length.out = 100),
  seed_diff = pred_means["seed_diff"]
)
grid_srs$yhat <- predict(bess_lm, newdata = grid_srs)

ggplot() +
  geom_jitter(data = train_set, aes(x = srs_diff, y = point_diff),
              alpha = 0.2, width = 0.1) +
  geom_line(data = grid_srs, aes(x = srs_diff, y = yhat),
            color = "steelblue", linewidth = 1.2) +
  labs(
    title    = "BeSS Regression: point_diff vs srs_diff",
    subtitle = "seed_diff held at its training-set mean",
    x = "SRS Difference (team1 - team2)",
    y = "Point Differential"
  ) +
  theme_bw()

# Plot 2: point_diff vs seed_diff (srs_diff held at its mean)
# seed_diff = seed.2 - seed.1: positive means team1 is better seeded (lower number)
grid_seed <- data.frame(
  srs_diff  = pred_means["srs_diff"],
  seed_diff = seq(quantile(train_set$seed_diff, 0.01),
                  quantile(train_set$seed_diff, 0.99), length.out = 100)
)
grid_seed$yhat <- predict(bess_lm, newdata = grid_seed)

ggplot() +
  geom_jitter(data = train_set, aes(x = seed_diff, y = point_diff),
              alpha = 0.2, width = 0.1) +
  geom_line(data = grid_seed, aes(x = seed_diff, y = yhat),
            color = "steelblue", linewidth = 1.2) +
  labs(
    title    = "BeSS Regression: point_diff vs seed_diff",
    subtitle = "srs_diff held at its training-set mean  |  positive seed_diff = team1 is better seeded",
    x = "Seed Difference (seed.2 - seed.1)",
    y = "Point Differential"
  ) +
  theme_bw()
```

```{r bess-reg-impact}
# Variable impact: predicted point_diff at 25th vs 75th percentile
# All other variables held at their training-set mean
impact_reg <- data.frame(variable = best_coefs_reg,
                         p25_pred = NA_real_,
                         p75_pred = NA_real_)

# For each var in best_coefs_reg, creates a base row where all vars 
# are set to their mean. Makes two copies, one with current var set to 25th% (low_row),
# and other set to its 75th% (high_row). Predicts the outcome (point_diff) for
# both rows using the fitted model (bess_lm). The stores predictions

for (i in seq_along(best_coefs_reg)) {
  var      <- best_coefs_reg[i]
  base     <- as.data.frame(t(pred_means))

  low_row        <- base; low_row[[var]]  <- quantile(train_set[[var]], 0.25)
  high_row       <- base; high_row[[var]] <- quantile(train_set[[var]], 0.75)

  impact_reg$p25_pred[i] <- predict(bess_lm, newdata = low_row)
  impact_reg$p75_pred[i] <- predict(bess_lm, newdata = high_row)
}

impact_reg$iqr_effect <- impact_reg$p75_pred - impact_reg$p25_pred
impact_reg <- impact_reg[order(abs(impact_reg$iqr_effect), decreasing = TRUE), ]
print(impact_reg)
```

```{r bess-reg-matchups}
# Helper: enrich any matchup dataframe with all model features
prepare_matchups <- function(df) {
  df %>%
    mutate(
      def_rtg.1    = (pts_pg_opp.1 / pace.1) * 100,
      def_rtg.2    = (pts_pg_opp.2 / pace.2) * 100,
      net_rtg.1    = off_rtg.1 - def_rtg.1,
      net_rtg.2    = off_rtg.2 - def_rtg.2,
      in_trap.1    = dist_to_trap(pace.1, net_rtg.1) <= 0,
      in_trap.2    = dist_to_trap(pace.2, net_rtg.2) <= 0,
      dist_trap.1  = dist_to_trap(pace.1, net_rtg.1),
      dist_trap.2  = dist_to_trap(pace.2, net_rtg.2),
      trap_advantage = case_when(
        in_trap.1 & !in_trap.2  ~  dist_trap.2,
        !in_trap.1 & in_trap.2  ~ -dist_trap.1,
        in_trap.1 & in_trap.2   ~  net_rtg.1 - net_rtg.2,
        TRUE                    ~  dist_trap.2 - dist_trap.1
      ),
      srs_diff     = srs.1     - srs.2,
      seed_diff    = seed.1    - seed.2,
      ts_pct_diff  = ts_pct.1  - ts_pct.2,
      tov_pct_diff = tov_pct.1 - tov_pct.2,
      trb_pct_diff = trb_pct.1 - trb_pct.2
    )
}

mens_matchups   <- prepare_matchups(mens_data[c(1509, 1511, 2613), ])
womens_matchups <- prepare_matchups(womens_data[c(1446, 1503, 1509), ])

mens_matchups$bess_pred_diff   <- predict(bess_lm, newdata = mens_matchups)
womens_matchups$bess_pred_diff <- predict(bess_lm, newdata = womens_matchups)

cat("--- Men's Matchups ---\n")
mens_matchups %>%
  select(name.1, name.2, seed.1, seed.2, srs_diff, trap_advantage, bess_pred_diff)

cat("--- Women's Matchups ---\n")
womens_matchups %>%
  select(name.1, name.2, seed.1, seed.2, srs_diff, trap_advantage, bess_pred_diff)
```

### BeSS Classification

```{r bess-class-fit}
# bess() with family = "binomial" for logistic best subset selection
# Response must be 0/1 integer
x_train_cl <- model.matrix(~ trap_advantage + srs_diff + seed_diff +
                              ts_pct_diff + tov_pct_diff + trb_pct_diff,
                            data = train_set)[, -1]  # drop intercept
x_test_cl  <- model.matrix(~ trap_advantage + srs_diff + seed_diff +
                              ts_pct_diff + tov_pct_diff + trb_pct_diff,
                            data = test_set)[, -1]   # drop intercept

y_train_cl <- as.integer(train_set$winner == "team1")  # 1 = team1 wins, 0 = team2 wins

bess_class <- bess(x = x_train_cl, y = y_train_cl,
                   family = "binomial", method = "sequential")

bess_class_summary <- data.frame(print(bess_class))

# Plot AIC and BIC vs model size
ggplot(bess_class_summary) +
  geom_line(aes(x = Df, y = AIC), linetype = "solid") +
  geom_line(aes(x = Df, y = BIC), linetype = "dashed") +
  annotate("point", x = which.min(bess_class_summary$AIC), y = min(bess_class_summary$AIC)) +
  annotate("point", x = which.min(bess_class_summary$BIC), y = min(bess_class_summary$BIC)) +
  labs(title = "BeSS Classification: AIC (solid) and BIC (dashed) vs Model Size",
       x = "Model Size", y = "AIC (solid) / BIC (dashed)")

# Best model size by BIC and its non-zero coefficients
best_size_class <- which.min(bess_class_summary$BIC)
cat("Best classification model size (BIC):", best_size_class, "\n")
BIC_coef_class <- coef(bess_class)[, best_size_class]
print(BIC_coef_class[BIC_coef_class != 0])
```

```{r bess-class-predict}
# Identify BIC-selected variables
best_coefs_class <- names(BIC_coef_class[BIC_coef_class != 0])
best_coefs_class <- best_coefs_class[!grepl("intercept", best_coefs_class, ignore.case = TRUE)]

# Refit as plain glm() for probabilities, accuracy, AUC, visualization, and impact analysis
bess_glm <- glm(as.formula(paste("I(winner == 'team1') ~",
                                  paste(best_coefs_class, collapse = " + "))),
                data = train_set, family = binomial)
summary(bess_glm)

# Predict probabilities on test set using the glm refit
prob_bess_class <- predict(bess_glm, newdata = test_set, type = "response")
pred_bess_class <- factor(ifelse(prob_bess_class >= 0.5, "team1", "team2"),
                          levels = c("team1", "team2"))

cm_bess <- confusionMatrix(pred_bess_class, test_set$winner, positive = "team1")
print(cm_bess)
cat("BeSS Classification — Accuracy:", round(cm_bess$overall["Accuracy"], 4), "\n")

roc_bess <- roc(test_set$winner, prob_bess_class, levels = c("team2", "team1"))
cat("BeSS Classification — AUC:", round(auc(roc_bess), 3), "\n")
plot(roc_bess, main = "ROC – BeSS Logistic Model", print.auc = TRUE)
```

```{r bess-class-viz}
# Plot P(team1 wins) vs each selected predictor, others held at training-set mean
class_means <- colMeans(train_set[, best_coefs_class, drop = FALSE])

for (var in best_coefs_class) {
  grid <- as.data.frame(
    lapply(best_coefs_class, function(v) {
      if (v == var) seq(quantile(train_set[[v]], 0.01),
                        quantile(train_set[[v]], 0.99), length.out = 100)
      else rep(class_means[v], 100)
    })
  )
  names(grid) <- best_coefs_class
  grid$p_team1 <- predict(bess_glm, newdata = grid, type = "response")

  print(
    ggplot(grid, aes(x = .data[[var]], y = p_team1)) +
      geom_line(color = "steelblue", linewidth = 1.2) +
      geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray40") +
      labs(
        title    = paste("BeSS Classification: P(team1 wins) vs", var),
        subtitle = "Other predictors held at their training-set mean",
        x = var, y = "P(team1 wins)"
      ) +
      theme_bw()
  )
}
```

```{r bess-class-impact}
# Variable impact: change in P(team1 wins) from 25th to 75th percentile
impact_class <- data.frame(variable = best_coefs_class,
                           p25_prob = NA_real_,
                           p75_prob = NA_real_)

for (i in seq_along(best_coefs_class)) {
  var      <- best_coefs_class[i]
  base     <- as.data.frame(t(class_means))

  low_row        <- base; low_row[[var]]  <- quantile(train_set[[var]], 0.25)
  high_row       <- base; high_row[[var]] <- quantile(train_set[[var]], 0.75)

  impact_class$p25_prob[i] <- predict(bess_glm, newdata = low_row,  type = "response")
  impact_class$p75_prob[i] <- predict(bess_glm, newdata = high_row, type = "response")
}

impact_class$iqr_effect <- impact_class$p75_prob - impact_class$p25_prob
impact_class <- impact_class[order(abs(impact_class$iqr_effect), decreasing = TRUE), ]
print(impact_class)
```

```{r bess-class-matchups}
mens_matchups$bess_p_team1   <- predict(bess_glm, newdata = mens_matchups,   type = "response")
womens_matchups$bess_p_team1 <- predict(bess_glm, newdata = womens_matchups, type = "response")

cat("--- Men's Matchups ---\n")
mens_matchups %>%
  select(name.1, name.2, seed.1, seed.2,
    srs_diff, seed_diff, bess_pred_diff, bess_p_team1)

cat("--- Women's Matchups ---\n")
womens_matchups %>%
  select(name.1, name.2, seed.1, seed.2,
    srs_diff, seed_diff, bess_pred_diff, bess_p_team1)
```

## LASSO

LASSO adds a penalty on the sum of the absolute values of the coefficients (the L1 norm). As the penalty strength λ increases, coefficients are shrunk continuously toward zero — some reach exactly zero and are effectively removed. We use `cv.glmnet()` to pick the best λ via cross-validation and report results at `lambda.1se`, the most parsimonious model within one standard error of the best CV error.

### LASSO Regression

```{r lasso-reg-fit}
library(glmnet)

# glmnet needs a numeric matrix — reuse the same model.matrix columns as BeSS
x_train_lasso_reg <- model.matrix(point_diff ~ trap_advantage + srs_diff + seed_diff +
                                    ts_pct_diff + tov_pct_diff + trb_pct_diff,
                                  data = train_set)[, -1]   # drop intercept column
x_test_lasso_reg  <- model.matrix(point_diff ~ trap_advantage + srs_diff + seed_diff +
                                    ts_pct_diff + tov_pct_diff + trb_pct_diff,
                                  data = test_set)[, -1]

# alpha = 1 -> LASSO (alpha = 0 would be Ridge)
set.seed(42)
lasso_reg_cv <- cv.glmnet(x_train_lasso_reg, train_set$point_diff,
                           alpha = 1, family = "gaussian")

# Plot CV error vs log(lambda) — vertical lines mark lambda.min and lambda.1se
plot(lasso_reg_cv)
title("LASSO Regression: CV Error vs log(Lambda)", line = 2.5)

cat("lambda.min :", round(lasso_reg_cv$lambda.min, 4), "\n")
cat("lambda.1se :", round(lasso_reg_cv$lambda.1se, 4), "\n")

# Coefficients at lambda.1se (most parsimonious model within 1 SE of best)
lasso_reg_coef <- coef(lasso_reg_cv, s = "lambda.1se")
print(lasso_reg_coef)

# Coefficient path across all lambda values
plot(lasso_reg_cv$glmnet.fit, xvar = "lambda", label = TRUE)
title("LASSO Regression: Coefficient Paths", line = 2.5)
```

```{r lasso-reg-predict}
# Test set predictions at lambda.1se
yhat_lasso_reg <- predict(lasso_reg_cv, newx = x_test_lasso_reg, s = "lambda.1se")
rmse_lasso_reg <- sqrt(mean((test_set$point_diff - yhat_lasso_reg)^2))
r2_lasso_reg   <- cor(test_set$point_diff, yhat_lasso_reg)^2
cat("LASSO Regression — Test RMSE:", round(rmse_lasso_reg, 3),
    " | Test R²:", round(r2_lasso_reg, 3), "\n")

# Extract non-zero variables for downstream use
lasso_reg_vars <- rownames(lasso_reg_coef)[as.vector(lasso_reg_coef) != 0]
lasso_reg_vars <- lasso_reg_vars[lasso_reg_vars != "(Intercept)"]
cat("Variables kept by LASSO (lambda.1se):", paste(lasso_reg_vars, collapse = ", "), "\n")

# Refit as plain lm() with surviving variables — needed for viz and impact
lasso_lm <- lm(as.formula(paste("point_diff ~", paste(lasso_reg_vars, collapse = " + "))),
               data = train_set)
summary(lasso_lm)
```

```{r lasso-reg-viz}
# Plot point_diff vs each surviving predictor, others held at training-set mean
lasso_reg_means <- colMeans(train_set[, lasso_reg_vars, drop = FALSE])

for (var in lasso_reg_vars) {
  grid <- as.data.frame(
    lapply(lasso_reg_vars, function(v) {
      if (v == var) seq(quantile(train_set[[v]], 0.01),
                        quantile(train_set[[v]], 0.99), length.out = 100)
      else rep(lasso_reg_means[v], 100)
    })
  )
  names(grid) <- lasso_reg_vars
  grid$yhat   <- predict(lasso_lm, newdata = grid)

  print(
    ggplot() +
      geom_jitter(data = train_set, aes(x = .data[[var]], y = point_diff),
                  alpha = 0.2, width = 0.1) +
      geom_line(data = grid, aes(x = .data[[var]], y = yhat),
                color = "darkorange", linewidth = 1.2) +
      labs(
        title    = paste("LASSO Regression: point_diff vs", var),
        subtitle = "Other predictors held at their training-set mean",
        x = var, y = "Point Differential"
      ) +
      theme_bw()
  )
}
```

```{r lasso-reg-impact}
# Variable impact: predicted point_diff at 25th vs 75th percentile
impact_lasso_reg <- data.frame(variable  = lasso_reg_vars,
                               p25_pred  = NA_real_,
                               p75_pred  = NA_real_)

for (i in seq_along(lasso_reg_vars)) {
  var      <- lasso_reg_vars[i]
  base     <- as.data.frame(t(lasso_reg_means))

  low_row        <- base; low_row[[var]]  <- quantile(train_set[[var]], 0.25)
  high_row       <- base; high_row[[var]] <- quantile(train_set[[var]], 0.75)

  impact_lasso_reg$p25_pred[i] <- predict(lasso_lm, newdata = low_row)
  impact_lasso_reg$p75_pred[i] <- predict(lasso_lm, newdata = high_row)
}

impact_lasso_reg$iqr_effect <- impact_lasso_reg$p75_pred - impact_lasso_reg$p25_pred
impact_lasso_reg <- impact_lasso_reg[order(abs(impact_lasso_reg$iqr_effect), decreasing = TRUE), ]
print(impact_lasso_reg)
```

```{r lasso-reg-matchups}
mens_matchups$lasso_pred_diff   <- predict(lasso_lm, newdata = mens_matchups)
womens_matchups$lasso_pred_diff <- predict(lasso_lm, newdata = womens_matchups)

cat("--- Men's Matchups ---\n")
mens_matchups %>%
  select(name.1, name.2, seed.1, seed.2,
         srs_diff, trap_advantage, bess_pred_diff, lasso_pred_diff)

cat("--- Women's Matchups ---\n")
womens_matchups %>%
  select(name.1, name.2, seed.1, seed.2,
         srs_diff, trap_advantage, bess_pred_diff, lasso_pred_diff)
```

### LASSO Classification

```{r lasso-class-fit}
# Response must be 0/1 for glmnet binomial
x_train_lasso_cl <- model.matrix(~ trap_advantage + srs_diff + seed_diff +
                                    ts_pct_diff + tov_pct_diff + trb_pct_diff,
                                  data = train_set)[, -1]
x_test_lasso_cl  <- model.matrix(~ trap_advantage + srs_diff + seed_diff +
                                    ts_pct_diff + tov_pct_diff + trb_pct_diff,
                                  data = test_set)[, -1]

y_train_lasso_cl <- as.integer(train_set$winner == "team1")

set.seed(42)
lasso_class_cv <- cv.glmnet(x_train_lasso_cl, y_train_lasso_cl,
                             alpha = 1, family = "binomial")

plot(lasso_class_cv)
title("LASSO Classification: CV Deviance vs log(Lambda)", line = 2.5)

cat("lambda.min :", round(lasso_class_cv$lambda.min, 4), "\n")
cat("lambda.1se :", round(lasso_class_cv$lambda.1se, 4), "\n")

lasso_class_coef <- coef(lasso_class_cv, s = "lambda.1se")
print(lasso_class_coef)

plot(lasso_class_cv$glmnet.fit, xvar = "lambda", label = TRUE)
title("LASSO Classification: Coefficient Paths", line = 2.5)
```

```{r lasso-class-predict}
# Predicted probabilities at lambda.1se
prob_lasso_class <- predict(lasso_class_cv, newx = x_test_lasso_cl,
                             s = "lambda.1se", type = "response")
pred_lasso_class <- factor(ifelse(prob_lasso_class >= 0.5, "team1", "team2"),
                            levels = c("team1", "team2"))

cm_lasso <- confusionMatrix(pred_lasso_class, test_set$winner, positive = "team1")
print(cm_lasso)

roc_lasso <- roc(test_set$winner, as.vector(prob_lasso_class),
                 levels = c("team2", "team1"))
cat("LASSO Classification — AUC:", round(auc(roc_lasso), 3), "\n")
plot(roc_lasso, main = "ROC – LASSO Logistic Model", print.auc = TRUE)

# Extract surviving variables and refit as plain glm() for viz/impact
lasso_class_vars <- rownames(lasso_class_coef)[as.vector(lasso_class_coef) != 0]
lasso_class_vars <- lasso_class_vars[lasso_class_vars != "(Intercept)"]
cat("Variables kept by LASSO (lambda.1se):", paste(lasso_class_vars, collapse = ", "), "\n")

lasso_glm <- glm(as.formula(paste("I(winner == 'team1') ~",
                                   paste(lasso_class_vars, collapse = " + "))),
                 data = train_set, family = binomial)
summary(lasso_glm)
```

```{r lasso-class-viz}
# Plot P(team1 wins) vs each surviving predictor, others at training-set mean
lasso_class_means <- colMeans(train_set[, lasso_class_vars, drop = FALSE])

for (var in lasso_class_vars) {
  grid <- as.data.frame(
    lapply(lasso_class_vars, function(v) {
      if (v == var) seq(quantile(train_set[[v]], 0.01),
                        quantile(train_set[[v]], 0.99), length.out = 100)
      else rep(lasso_class_means[v], 100)
    })
  )
  names(grid) <- lasso_class_vars
  grid$p_team1 <- predict(lasso_glm, newdata = grid, type = "response")

  print(
    ggplot(grid, aes(x = .data[[var]], y = p_team1)) +
      geom_line(color = "darkorange", linewidth = 1.2) +
      geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray40") +
      labs(
        title    = paste("LASSO Classification: P(team1 wins) vs", var),
        subtitle = "Other predictors held at their training-set mean",
        x = var, y = "P(team1 wins)"
      ) +
      theme_bw()
  )
}
```

```{r lasso-class-impact}
# Variable impact: change in P(team1 wins) from 25th to 75th percentile
impact_lasso_class <- data.frame(variable = lasso_class_vars,
                                 p25_prob = NA_real_,
                                 p75_prob = NA_real_)

for (i in seq_along(lasso_class_vars)) {
  var      <- lasso_class_vars[i]
  base     <- as.data.frame(t(lasso_class_means))

  low_row        <- base; low_row[[var]]  <- quantile(train_set[[var]], 0.25)
  high_row       <- base; high_row[[var]] <- quantile(train_set[[var]], 0.75)

  impact_lasso_class$p25_prob[i] <- predict(lasso_glm, newdata = low_row,  type = "response")
  impact_lasso_class$p75_prob[i] <- predict(lasso_glm, newdata = high_row, type = "response")
}

impact_lasso_class$iqr_effect <- impact_lasso_class$p75_prob - impact_lasso_class$p25_prob
impact_lasso_class <- impact_lasso_class[order(abs(impact_lasso_class$iqr_effect), decreasing = TRUE), ]
print(impact_lasso_class)
```

```{r lasso-class-matchups}
mens_matchups$lasso_p_team1   <- predict(lasso_glm, newdata = mens_matchups,   type = "response")
womens_matchups$lasso_p_team1 <- predict(lasso_glm, newdata = womens_matchups, type = "response")

cat("--- Men's Matchups ---\n")
mens_matchups %>%
  select(name.1, name.2, seed.1, seed.2,
         srs_diff, seed_diff,
         bess_p_team1, lasso_p_team1)

cat("--- Women's Matchups ---\n")
womens_matchups %>%
  select(name.1, name.2, seed.1, seed.2,
         srs_diff, seed_diff,
         bess_p_team1, lasso_p_team1)
```

## k-Nearest Neighbors (kNN)

kNN makes predictions by finding the k most similar training observations (neighbors) to a new data point and averaging their outcomes (regression) or taking a majority vote (classification). Because kNN uses Euclidean distance, all predictors must be **centered and scaled** first — otherwise a variable with a large numeric range would dominate the distance calculation unfairly. We use `caret::train()` with cross-validation to automatically select the best k.
I. 
### kNN Regression

```{r knn-reg-fit}
# All 6 predictors — caret will center+scale internally via preProcess
knn_predictors <- c("trap_advantage", "srs_diff", "seed_diff",
                    "ts_pct_diff", "tov_pct_diff", "trb_pct_diff")

# 10-fold CV (class standard) to choose the best k
ctrl_reg <- trainControl(method = "cv", number = 10)

set.seed(42)
knn_reg <- train(
  x          = train_set[, knn_predictors],
  y          = train_set$point_diff,
  method     = "knn",
  preProcess = c("center", "scale"),
  trControl  = ctrl_reg,
  tuneGrid   = data.frame(k = seq(1, 41, 2))  # odd k values 1–41
)

print(knn_reg)

# Plot CV RMSE vs k
ggplot(knn_reg) +
  labs(title = "kNN Regression: CV RMSE vs k",
       x = "k (number of neighbors)", y = "Cross-Validated RMSE") +
  theme_bw()

cat("Best k (regression):", knn_reg$bestTune$k, "\n")
```

```{r knn-reg-predict}
# Test set performance
yhat_knn_reg   <- predict(knn_reg, newdata = test_set[, knn_predictors])
rmse_knn_reg   <- sqrt(mean((test_set$point_diff - yhat_knn_reg)^2))
r2_knn_reg     <- cor(test_set$point_diff, yhat_knn_reg)^2
cat("kNN Regression — Test RMSE:", round(rmse_knn_reg, 3),
    " | Test R²:", round(r2_knn_reg, 3), "\n")
```

```{r knn-reg-viz}
# Partial effect plots: vary one predictor, hold others at training-set mean
knn_reg_means <- colMeans(train_set[, knn_predictors])

for (var in knn_predictors) {
  grid <- as.data.frame(
    lapply(knn_predictors, function(v) {
      if (v == var) seq(quantile(train_set[[v]], 0.01),
                        quantile(train_set[[v]], 0.99), length.out = 100)
      else rep(knn_reg_means[v], 100)
    })
  )
  names(grid) <- knn_predictors
  grid$yhat   <- predict(knn_reg, newdata = grid)

  print(
    ggplot() +
      geom_jitter(data = train_set, aes(x = .data[[var]], y = point_diff),
                  alpha = 0.2, width = 0.1) +
      geom_line(data = grid, aes(x = .data[[var]], y = yhat),
                color = "purple", linewidth = 1.2) +
      labs(
        title    = paste("kNN Regression: point_diff vs", var),
        subtitle = paste0("k = ", knn_reg$bestTune$k,
                          "; other predictors held at training-set mean"),
        x = var, y = "Point Differential"
      ) +
      theme_bw()
  )
}
```

```{r knn-reg-impact}
# Variable impact: predicted point_diff at 25th vs 75th percentile
impact_knn_reg <- data.frame(variable = knn_predictors,
                             p25_pred = NA_real_,
                             p75_pred = NA_real_)

for (i in seq_along(knn_predictors)) {
  var  <- knn_predictors[i]
  base <- as.data.frame(t(knn_reg_means))

  low_row        <- base; low_row[[var]]  <- quantile(train_set[[var]], 0.25)
  high_row       <- base; high_row[[var]] <- quantile(train_set[[var]], 0.75)

  impact_knn_reg$p25_pred[i] <- predict(knn_reg, newdata = low_row)
  impact_knn_reg$p75_pred[i] <- predict(knn_reg, newdata = high_row)
}

impact_knn_reg$iqr_effect <- impact_knn_reg$p75_pred - impact_knn_reg$p25_pred
impact_knn_reg <- impact_knn_reg[order(abs(impact_knn_reg$iqr_effect), decreasing = TRUE), ]
print(impact_knn_reg)
```

```{r knn-reg-matchups}
mens_matchups$knn_pred_diff   <- predict(knn_reg, newdata = mens_matchups[,   knn_predictors])
womens_matchups$knn_pred_diff <- predict(knn_reg, newdata = womens_matchups[, knn_predictors])

cat("--- Men's Matchups ---\n")
mens_matchups %>%
  select(name.1, name.2, seed.1, seed.2,
         srs_diff, trap_advantage,
         bess_pred_diff, lasso_pred_diff, knn_pred_diff)

cat("--- Women's Matchups ---\n")
womens_matchups %>%
  select(name.1, name.2, seed.1, seed.2,
         srs_diff, trap_advantage,
         bess_pred_diff, lasso_pred_diff, knn_pred_diff)
```

### kNN Classification

```{r knn-class-fit}
# 10-fold CV, use Kappa as selection metric (class standard for classification)
ctrl_class <- trainControl(method = "cv", number = 10)

set.seed(42)
knn_class <- train(
  x          = train_set[, knn_predictors],
  y          = train_set$winner,   # factor with levels "team1", "team2"
  method     = "knn",
  preProcess = c("center", "scale"),
  trControl  = ctrl_class,
  tuneGrid   = data.frame(k = seq(1, 41, 2)),  # odd k values 1–41
  metric     = "Kappa"             # select best k by Cohen's Kappa (class standard)
)

print(knn_class)

# Plot CV AUC vs k
ggplot(knn_class) +
  labs(title = "kNN Classification: CV AUC vs k",
       x = "k (number of neighbors)", y = "Cross-Validated AUC") +
  theme_bw()

cat("Best k (classification):", knn_class$bestTune$k, "\n")
```

```{r knn-class-predict}
# Predicted probabilities and class labels on test set
prob_knn_class <- predict(knn_class, newdata = test_set[, knn_predictors], type = "prob")
pred_knn_class <- predict(knn_class, newdata = test_set[, knn_predictors])

cm_knn <- confusionMatrix(pred_knn_class, test_set$winner, positive = "team1")
print(cm_knn)

roc_knn <- roc(test_set$winner, prob_knn_class$team1, levels = c("team2", "team1"))
cat("kNN Classification — AUC:", round(auc(roc_knn), 3), "\n")
plot(roc_knn, main = "ROC – kNN Classification Model", print.auc = TRUE)
```

```{r knn-class-viz}
# Partial effect plots: vary one predictor, hold others at training-set mean
knn_class_means <- colMeans(train_set[, knn_predictors])

for (var in knn_predictors) {
  grid <- as.data.frame(
    lapply(knn_predictors, function(v) {
      if (v == var) seq(quantile(train_set[[v]], 0.01),
                        quantile(train_set[[v]], 0.99), length.out = 100)
      else rep(knn_class_means[v], 100)
    })
  )
  names(grid) <- knn_predictors
  grid$p_team1 <- predict(knn_class, newdata = grid, type = "prob")$team1

  print(
    ggplot(grid, aes(x = .data[[var]], y = p_team1)) +
      geom_line(color = "purple", linewidth = 1.2) +
      geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray40") +
      labs(
        title    = paste("kNN Classification: P(team1 wins) vs", var),
        subtitle = paste0("k = ", knn_class$bestTune$k,
                          "; other predictors held at training-set mean"),
        x = var, y = "P(team1 wins)"
      ) +
      theme_bw()
  )
}
```

```{r knn-class-impact}
# Variable impact: change in P(team1 wins) from 25th to 75th percentile
impact_knn_class <- data.frame(variable = knn_predictors,
                               p25_prob = NA_real_,
                               p75_prob = NA_real_)

for (i in seq_along(knn_predictors)) {
  var  <- knn_predictors[i]
  base <- as.data.frame(t(knn_class_means))

  low_row        <- base; low_row[[var]]  <- quantile(train_set[[var]], 0.25)
  high_row       <- base; high_row[[var]] <- quantile(train_set[[var]], 0.75)

  impact_knn_class$p25_prob[i] <- predict(knn_class, newdata = low_row,  type = "prob")$team1
  impact_knn_class$p75_prob[i] <- predict(knn_class, newdata = high_row, type = "prob")$team1
}

impact_knn_class$iqr_effect <- impact_knn_class$p75_prob - impact_knn_class$p25_prob
impact_knn_class <- impact_knn_class[order(abs(impact_knn_class$iqr_effect), decreasing = TRUE), ]
print(impact_knn_class)
```

```{r knn-class-matchups}
mens_matchups$knn_p_team1   <- predict(knn_class, newdata = mens_matchups[,   knn_predictors], type = "prob")$team1
womens_matchups$knn_p_team1 <- predict(knn_class, newdata = womens_matchups[, knn_predictors], type = "prob")$team1

cat("--- Men's Matchups ---\n")
mens_matchups %>%
  select(name.1, name.2, seed.1, seed.2,
         srs_diff, seed_diff,
         bess_p_team1, lasso_p_team1, knn_p_team1)

cat("--- Women's Matchups ---\n")
womens_matchups %>%
  select(name.1, name.2, seed.1, seed.2,
         srs_diff, seed_diff,
         bess_p_team1, lasso_p_team1, knn_p_team1)
```
