---
title: "MIAC Baseball 2026 - Full Season Scraper"
author: "Ben Trutwin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
```

## Overview

This script scrapes all MIAC baseball data for the 2026 season end-to-end with no manual intervention:

1. Scrapes the full composite schedule (all games, all months)
2. Scrapes every box score for every completed game
3. Scrapes team and individual cumulative stats from the MIAC stats page
4. Cleans and merges all data into a unified dataset
5. Saves all outputs as CSV files

---

## Libraries and Constants

> **Note on `chromote`:** The MIAC calendar page renders its schedule table via JavaScript, so plain `rvest::read_html()` returns an empty shell with no game rows. `chromote` launches a headless Chrome browser from R, executes the JavaScript, and then hands the fully-rendered HTML back to `rvest` for parsing. You need Google Chrome installed; then run `install.packages("chromote")` once.

```{r libraries-constants}
library(rvest)
library(dplyr)
library(stringr)
library(purrr)
library(tidyr)
library(janitor)
library(chromote)   # headless Chrome for JavaScript-rendered pages

BASE_URL     <- "https://miacathletics.com"
CALENDAR_URL <- paste0(BASE_URL, "/calendar.aspx?path=baseball")
STATS_URL    <- paste0(BASE_URL, "/stats.aspx?path=baseball&year=2026")
YEAR         <- 2026
```

---

## Function Definitions

### `scrape_calendar_month()`

Scrapes the MIAC baseball calendar for a given year and month using a headless Chrome browser (`chromote`) to execute the JavaScript that populates the schedule table before parsing. **Note:** The MIAC composite schedule is not actually paginated by month — every month URL returns the full season schedule (Feb–May 2026). This function exists for modularity and robustness.

```{r fn-scrape-calendar-month}
scrape_calendar_month <- function(year, month) {
  url <- paste0(BASE_URL, "/calendar.aspx?path=baseball&year=", year, "&month=", month)

  # Launch a headless Chrome session to execute JavaScript before parsing.
  # The schedule table is populated by JS, so plain read_html() returns no rows.
  html_content <- tryCatch({
    b <- ChromoteSession$new()
    on.exit(b$close(), add = TRUE)   # always close the session, even on error

    b$Page$navigate(url, wait_ = FALSE)
    b$Page$loadEventFired()          # wait for the initial page load
    Sys.sleep(2)                     # extra buffer for JS table population

    b$Runtime$evaluate('document.documentElement.outerHTML')$result$value
  }, error = function(e) {
    message("  [WARN] chromote failed for month ", month, ": ", e$message)
    return(NULL)
  })

  if (is.null(html_content)) return(NULL)

  page <- read_html(html_content)
  extract_results_and_links(page)
}
```

### `extract_results_and_links()`

Parses the raw HTML of a calendar page to extract game rows. Each row contains: date, away team + score, home team + score, result, location, and box score URL. For cancelled/postponed games, scores are absent and no box score link exists.

The MIAC schedule table structure per row:

`| Date | AwayTeam Score | HomeTeam Score | Result | Location | BoxScore |`

```{r fn-extract-results-and-links}
extract_results_and_links <- function(page_html) {

  rows <- page_html %>% html_nodes("table tbody tr")

  if (length(rows) == 0) {
    message("  [INFO] No table rows found on this page.")
    return(tibble(
      date = character(), away_raw = character(), home_raw = character(),
      result = character(), location = character(), boxscore_url = character()
    ))
  }

  row_data <- map(rows, function(r) {
    cells <- r %>% html_nodes("td") %>% html_text(trim = TRUE)

    if (length(cells) < 4) return(NULL)
    if (all(cells == "" | cells == " ")) return(NULL)

    hrefs    <- r %>% html_nodes("a") %>% html_attr("href")
    bs_hrefs <- hrefs[str_detect(hrefs, regex("boxscore", ignore_case = TRUE))]

    boxscore_url <- NA_character_
    if (length(bs_hrefs) > 0) {
      bs <- bs_hrefs[1]
      boxscore_url <- ifelse(str_starts(bs, "http"), bs, paste0(BASE_URL, bs))
    }

    # cells[1] = date, cells[2] = away team+score, cells[3] = home team+score
    # cells[4] = result (e.g. "Final 9 innings"), cells[5] = location
    date_val     <- if (length(cells) >= 1) cells[1] else NA_character_
    away_raw     <- if (length(cells) >= 2) cells[2] else NA_character_
    home_raw     <- if (length(cells) >= 3) cells[3] else NA_character_
    result_val   <- if (length(cells) >= 4) cells[4] else NA_character_
    location_val <- if (length(cells) >= 5) cells[5] else NA_character_

    if (is.na(date_val) || date_val == "" || !str_detect(date_val, "\\d{1,2}/\\d{1,2}/\\d{4}")) {
      return(NULL)
    }

    tibble(
      date         = date_val,
      away_raw     = away_raw,
      home_raw     = home_raw,
      result       = result_val,
      location     = location_val,
      boxscore_url = boxscore_url
    )
  })

  row_data %>%
    compact() %>%
    bind_rows()
}
```

### `parse_team_score()`

Splits a combined cell like `"Hamline 4"` into a team name and an integer score. Handles cancelled/postponed games where no score is present.

```{r fn-parse-team-score}
parse_team_score <- function(raw_text) {
  if (is.na(raw_text) || raw_text == "") {
    return(list(team = NA_character_, score = NA_integer_))
  }

  # The calendar cells contain multi-line text with embedded "(Away)"/"(Home)"
  # role labels and lots of whitespace. Collapse and clean before matching.
  cleaned <- raw_text %>%
    str_replace_all("\\s*\\(Away\\)|\\s*\\(Home\\)", "") %>%  # strip role labels
    str_replace_all("[\\r\\n\\t]+", " ") %>%                  # collapse line breaks
    str_squish()                                               # normalise whitespace

  # Match the last number in the string as the score
  # e.g., "Massachusetts Institute of Technology 1" -> team + score
  # e.g., "Saint Mary's" (cancelled) -> team only
  match <- str_match(cleaned, "^(.+?)\\s+(\\d+)$")

  if (is.na(match[1, 1])) {
    return(list(team = str_trim(cleaned), score = NA_integer_))
  }

  list(
    team  = str_trim(match[1, 2]),
    score = as.integer(match[1, 3])
  )
}
```

### `scrape_boxscore()`

Scrapes a single box score page and extracts runs (from the linescore table) and starting pitcher info (from the pitching tables). Returns a single-row tibble with NA values on failure rather than crashing.

```{r fn-scrape-boxscore}
scrape_boxscore <- function(url) {
  na_row <- tibble(
    boxscore_url    = url,
    away_runs       = NA_integer_,
    home_runs       = NA_integer_,
    away_starter    = NA_character_,
    home_starter    = NA_character_,
    away_starter_ip = NA_character_,
    home_starter_ip = NA_character_
  )

  if (is.na(url) || url == "") return(na_row)

  pg <- tryCatch(read_html(url), error = function(e) NULL)
  if (is.null(pg)) {
    message("  [WARN] Could not load box score: ", url)
    return(na_row)
  }

  tbls <- tryCatch(pg %>% html_table(fill = TRUE), error = function(e) list())
  if (length(tbls) == 0) return(na_row)

  # --- LINESCORE: first table with an "R" (runs) column ---
  ls_idx    <- which(sapply(tbls, function(x) "R" %in% names(x)))
  away_runs <- NA_integer_
  home_runs <- NA_integer_

  if (length(ls_idx) > 0) {
    linescore <- tbls[[ls_idx[1]]]
    if (nrow(linescore) >= 2 && "R" %in% names(linescore)) {
      away_runs <- suppressWarnings(as.integer(linescore$R[1]))
      home_runs <- suppressWarnings(as.integer(linescore$R[2]))
    }
  }

  # --- PITCHING: tables with an "IP" column (away first, home second) ---
  pitching_tbls   <- tbls[sapply(tbls, function(x) "IP" %in% names(x))]
  away_starter    <- NA_character_
  home_starter    <- NA_character_
  away_starter_ip <- NA_character_
  home_starter_ip <- NA_character_

  find_player_col <- function(tbl) {
    col_names  <- names(tbl)
    player_col <- col_names[str_detect(col_names, regex("player|name", ignore_case = TRUE))]
    if (length(player_col) > 0) return(player_col[1])
    col_names[1]
  }

  if (length(pitching_tbls) >= 1) {
    pt1 <- pitching_tbls[[1]] %>%
      filter(!str_detect(.data[[find_player_col(pitching_tbls[[1]])]], regex("total|team", ignore_case = TRUE)))
    if (nrow(pt1) >= 1) {
      away_starter    <- as.character(pt1[[find_player_col(pt1)]][1])
      away_starter_ip <- as.character(pt1[["IP"]][1])
    }
  }

  if (length(pitching_tbls) >= 2) {
    pt2 <- pitching_tbls[[2]] %>%
      filter(!str_detect(.data[[find_player_col(pitching_tbls[[2]])]], regex("total|team", ignore_case = TRUE)))
    if (nrow(pt2) >= 1) {
      home_starter    <- as.character(pt2[[find_player_col(pt2)]][1])
      home_starter_ip <- as.character(pt2[["IP"]][1])
    }
  }

  tibble(
    boxscore_url    = url,
    away_runs       = away_runs,
    home_runs       = home_runs,
    away_starter    = away_starter,
    home_starter    = home_starter,
    away_starter_ip = away_starter_ip,
    home_starter_ip = home_starter_ip
  )
}
```

### `scrape_stats()`

Scrapes the MIAC cumulative stats page for a given year. Identifies tables by **column patterns** (not fixed indices) for robustness against layout changes. Returns a named list of tibbles for team batting, team pitching, team fielding, individual batting, and individual pitching.

```{r fn-scrape-stats}
scrape_stats <- function(year) {
  url <- paste0(BASE_URL, "/stats.aspx?path=baseball&year=", year)

  page <- tryCatch(read_html(url), error = function(e) {
    stop("Could not load stats page: ", e$message)
  })

  tbls <- page %>% html_table(fill = TRUE)
  message("  [INFO] Found ", length(tbls), " tables on stats page.")

  is_batting_table <- function(tbl) {
    cols <- toupper(names(tbl))
    all(c("AVG") %in% cols) && any(c("AB", "R", "H") %in% cols)
  }

  is_pitching_table <- function(tbl) {
    cols <- toupper(names(tbl))
    any(c("ERA") %in% cols) && any(c("IP", "W-L") %in% cols)
  }

  is_fielding_table <- function(tbl) {
    cols <- toupper(names(tbl))
    any(c("FLD%") %in% cols) && any(c("PO", "A", "E") %in% cols)
  }

  has_player_column <- function(tbl) {
    any(str_detect(toupper(names(tbl)), "PLAYER|NAME"))
  }

  team_batting_list   <- list()
  team_pitching_list  <- list()
  team_fielding_list  <- list()
  indiv_batting_list  <- list()
  indiv_pitching_list <- list()

  for (i in seq_along(tbls)) {
    tbl <- tbls[[i]]
    if (ncol(tbl) < 3 || nrow(tbl) < 2) next

    # Check batting BEFORE fielding: the team batting table includes PO/A/E/FLD%
    # columns, so is_fielding_table() would match it first and misclassify it.
    if (is_batting_table(tbl)) {
      if (has_player_column(tbl)) {
        indiv_batting_list <- append(indiv_batting_list, list(tbl))
      } else {
        team_batting_list <- append(team_batting_list, list(tbl))
      }
    } else if (is_pitching_table(tbl)) {
      if (has_player_column(tbl)) {
        indiv_pitching_list <- append(indiv_pitching_list, list(tbl))
      } else {
        team_pitching_list <- append(team_pitching_list, list(tbl))
      }
    } else if (is_fielding_table(tbl)) {
      if (!has_player_column(tbl)) {
        team_fielding_list <- append(team_fielding_list, list(tbl))
      }
    }
  }

  pick_first <- function(lst, label) {
    if (length(lst) > 0) {
      message("  [INFO] Found ", length(lst), " candidate(s) for ", label, "; using first.")
      return(lst[[1]])
    }
    message("  [WARN] No table found for ", label)
    return(tibble())
  }

  list(
    team_batting        = pick_first(team_batting_list,   "team_batting"),
    team_pitching       = pick_first(team_pitching_list,  "team_pitching"),
    team_fielding       = pick_first(team_fielding_list,  "team_fielding"),
    individual_batting  = pick_first(indiv_batting_list,  "individual_batting"),
    individual_pitching = pick_first(indiv_pitching_list, "individual_pitching")
  )
}
```

### `clean_games_table()`

Cleans the raw games data frame by:

- Parsing team names and scores from the combined raw cells
- Converting date strings to `Date` objects and extracting day of week
- Flagging conference games (location contains `"(Conf.)"`)
- Normalizing the result status to `Final`, `Cancelled`, or `Postponed`
- Determining the winner of each completed game

```{r fn-clean-games-table}
clean_games_table <- function(df) {
  if (nrow(df) == 0) return(df)

  away_parsed <- map(df$away_raw, parse_team_score)
  home_parsed <- map(df$home_raw, parse_team_score)

  df <- df %>%
    mutate(
      away_team  = map_chr(away_parsed, ~ .x$team),
      away_score = map_int(away_parsed, ~ .x$score),
      home_team  = map_chr(home_parsed, ~ .x$team),
      home_score = map_int(home_parsed, ~ .x$score),

      date_clean  = str_extract(date, "\\d{1,2}/\\d{1,2}/\\d{4}"),
      game_date   = as.Date(date_clean, format = "%m/%d/%Y"),
      day_of_week = weekdays(game_date),

      is_conference = str_detect(location, fixed("(Conf.)")),

      status = case_when(
        str_detect(result, regex("final",     ignore_case = TRUE)) ~ "Final",
        str_detect(result, regex("cancelled", ignore_case = TRUE)) ~ "Cancelled",
        str_detect(result, regex("postponed", ignore_case = TRUE)) ~ "Postponed",
        TRUE ~ result
      ),

      winner = case_when(
        is.na(away_score) | is.na(home_score) ~ NA_character_,
        away_score > home_score               ~ away_team,
        home_score > away_score               ~ home_team,
        TRUE                                  ~ "Tie"
      )
    ) %>%
    select(
      game_date, day_of_week, away_team, away_score, home_team, home_score,
      status, result, winner, is_conference, location, boxscore_url,
      everything(), -date_clean, -away_raw, -home_raw, -date
    )

  df
}
```

---

## Step 1: Scrape the Composite Schedule

The MIAC composite schedule is not paginated by month — every month URL returns the identical full-season schedule (Feb 14 – May 5, 2026). Because each call now spins up a headless Chrome browser, looping all 12 months would waste ~12 browser launches for duplicate data. Instead, we scrape the base calendar URL once and get everything in a single pass.

```{r step1-scrape-schedule}
cat("Scraping composite schedule...\n")

games_raw <- scrape_calendar_month(YEAR, month = 2)  # month=2 covers full season; any value works

if (is.null(games_raw) || nrow(games_raw) == 0) {
  stop("No games were returned. Check that Chrome is installed and chromote is working.")
}

cat("Total unique games found:", nrow(games_raw), "\n")
head(games_raw)
```

---

## Step 2: Clean the Games Table

```{r step2-clean-games}
games_clean <- clean_games_table(games_raw)
cat("Games after cleaning:", nrow(games_clean), "\n")
head(games_clean)
```

---

## Step 3: Scrape Box Scores

Only scrape completed games that have a box score URL.

```{r step3-scrape-boxscores}
urls_to_scrape <- games_clean %>%
  filter(!is.na(boxscore_url), status == "Final") %>%
  pull(boxscore_url) %>%
  unique()

cat("Box scores to scrape:", length(urls_to_scrape), "\n")

boxscores <- map_dfr(seq_along(urls_to_scrape), function(i) {
  if (i %% 10 == 0 || i == 1) {
    cat("  -> Scraping box score", i, "of", length(urls_to_scrape), "...\n")
  }
  result <- tryCatch(
    scrape_boxscore(urls_to_scrape[i]),
    error = function(e) {
      message("  [ERROR] Box score ", i, ": ", e$message)
      tibble(
        boxscore_url    = urls_to_scrape[i],
        away_runs       = NA_integer_,
        home_runs       = NA_integer_,
        away_starter    = NA_character_,
        home_starter    = NA_character_,
        away_starter_ip = NA_character_,
        home_starter_ip = NA_character_
      )
    }
  )
  Sys.sleep(0.5)  # Be polite
  result
})

cat("Box scores scraped:", nrow(boxscores), "\n")
head(boxscores)
```

---

## Step 4: Scrape Cumulative Stats

```{r step4-scrape-stats}
cat("Scraping cumulative stats...\n")
stats <- scrape_stats(YEAR)

team_batting        <- stats$team_batting
team_pitching       <- stats$team_pitching
team_fielding       <- stats$team_fielding
individual_batting  <- stats$individual_batting
individual_pitching <- stats$individual_pitching

cat("Team batting rows:       ", nrow(team_batting), "\n")
cat("Team pitching rows:      ", nrow(team_pitching), "\n")
cat("Team fielding rows:      ", nrow(team_fielding), "\n")
cat("Individual batting rows: ", nrow(individual_batting), "\n")
cat("Individual pitching rows:", nrow(individual_pitching), "\n")
```

---

## Step 5: Merge and Save CSVs

```{r step5-merge-save}
# Merge games with box score data
games_full <- games_clean %>%
  left_join(boxscores, by = "boxscore_url")

# Save all CSVs
write.csv(team_batting,        "team_batting_2026.csv",        row.names = FALSE)
write.csv(team_pitching,       "team_pitching_2026.csv",       row.names = FALSE)
write.csv(team_fielding,       "team_fielding_2026.csv",       row.names = FALSE)
write.csv(individual_batting,  "individual_batting_2026.csv",  row.names = FALSE)
write.csv(individual_pitching, "individual_pitching_2026.csv", row.names = FALSE)
write.csv(games_clean,         "miac_games_2026.csv",          row.names = FALSE)
write.csv(boxscores,           "miac_boxscores_2026.csv",      row.names = FALSE)
write.csv(games_full,          "miac_games_full_2026.csv",     row.names = FALSE)

cat("All CSVs saved:\n")
cat("  - team_batting_2026.csv\n")
cat("  - team_pitching_2026.csv\n")
cat("  - team_fielding_2026.csv\n")
cat("  - individual_batting_2026.csv\n")
cat("  - individual_pitching_2026.csv\n")
cat("  - miac_games_2026.csv\n")
cat("  - miac_boxscores_2026.csv\n")
cat("  - miac_games_full_2026.csv\n")
```

---

## Sanity Check

```{r sanity-check}
cat("games_full rows:", nrow(games_full), "\n")
head(games_full)

cat("\nteam_batting rows:", nrow(team_batting), "\n")
head(team_batting)

cat("\nindividual_pitching rows:", nrow(individual_pitching), "\n")
head(individual_pitching)
```
